Reproducible Research Peer Assessment 1
========================================================

This is an peer assesment 1 for Coursera course Reproducible Research from Data Science Specialization.

## Loading data

```{r loading.data}
dt <- read.csv("./data/activity.csv")
```

## Histogram of the total number of steps

```{r histogram, fig.width=7, fig.height=6}
dt.aggr <- aggregate(.~date, data=dt, sum)
hist(
    x=dt.aggr$steps, 
    main="Histogram of steps per day",
    xlab="Steps per day"
    )
```

## Mean and median of total number of steps
  
```{r mean.median}
s.mean <- as.integer(round(mean(dt.aggr$steps),0))
s.median <- median(dt.aggr$steps)
```
The mean of steps taken per day is `r s.mean`.  
The meadian of steps taken per day is `r s.median`.

## What is the average daily activity pattern?

Ploting average number of steps per day.

```{r steps.mean.plot}

dt.aggr.2 <- aggregate(.~interval, data=dt, mean)
time.int <- c("0:00", "3:00", "6:00", "9:00", "12:00",
              "15:00", "18:00", "21:00", "24:00")
plot(
    x=dt.aggr.2$interval, 
    y=dt.aggr.2$steps, 
    type="l", 
    axes=F,
    xlab="time interval",
    ylab="mean of steps",
    main="")
axis(1, at=c(0,300,600,900,1200,1500,1800,2100,2400), lab=time.int)
axis(2, at=c(0,50,100,150,200))

```

Calculating maximum number of steps on average across all the day.

```{r day.peak}
# finding peak
day.peak <- as.character(dt.aggr.2$interval[dt.aggr.2$steps==max(dt.aggr.2$steps)])

# formating peak to sutible format
# inserting : fird place from the end of string
day.peak <- sub( '(?<=.{1})', ':', day.peak, perl=TRUE )
```

On average maximum steps are made on `r day.peak`


## Missing values 
### Number of rows with missing value

```{r missing.val}
sum.na <- sum(is.na(dt$steps))
```

In the data set was found **`r sum.na`** mising value (rows with missing value)

### Missing values strategy

The missing values was filled with the mean for that 5-minute interval.

### Data set without NA

``` {r data.without.NA}
dt.no.na <- dt 

for(i in 1:dim(dt.no.na)[1]){
    if(is.na(dt.no.na$steps[i])){
        dt.no.na$steps[i] <- dt.aggr.2$steps[dt.aggr.2$interval==dt.no.na$interval[i]]
    }
}

```

### Histogram and summary statistic of the data without NA

Ploting original histogram of the steps per day and histogram with removed NA.

```{r hist.witout.NA, fig.width=8, fig.height=6}

dt2.aggr <- aggregate(.~date, data=dt.no.na, sum)

par(mfrow=c(1,2))
    hist(
        x=dt2.aggr$steps, 
        main="Histogram of steps per day 
        with removed NA",
        xlab="Steps per day"
    )

    hist(
        x=dt.aggr$steps, 
        main="Histogram of steps per day",
        xlab="Steps per day"
    )

par(mfrow=c(1,1))
```

```{r mean.and.med.without.NA}
step.mean.NA <- as.integer(mean(dt2.aggr$steps))
step.median.NA <- as.integer(median(dt2.aggr$steps))
```

Mean of steps taken per day if the missing values are filled with the mean for that 5-minute interval is **`r step.mean.NA`**

Median of steps taken per day if the missing values are filled with the mean for that 5-minute interval is **`r step.median.NA`**

By adding mean insted of NA we did not change the shape of the data per day.  

## Differences in activity patterns between weekdays and weekends.
  
Ploting average steps on weekend and on workday.
  
```{r weekends, fig.width=8, fig.height=6}

# making colummn with weekday or weekend as factor
dt.no.na$days <- NA
for(i in 1:dim(dt.no.na)[1]){
    if(weekdays(as.Date(dt.no.na$date[i])) == "Sunday" | 
       weekdays(as.Date(dt.no.na$date[i])) == "Saturday"){
        dt.no.na$days[i] <- "Weekend"
    } else {
        dt.no.na$days[i] <- "Weekday"
    }
}
dt.no.na$days <- as.factor(dt.no.na$days)

# calculating mean for eevery 5 min for weekday and weekend separately

aggr.days <- aggregate(.~interval, 
                       data=subset(x=dt.no.na, subset=dt.no.na$days == "Weekday"), 
                       mean)
aggr.days$days <- "Weekdays"

aggr.ends <- aggregate(.~interval, 
                       data=subset(x=dt.no.na, subset=dt.no.na$days == "Weekend"), 
                       mean)
aggr.ends$days <- "Weekends"

# adding all together 

aggr.melt <- data.frame(interval = rep(NA, times=576), steps=NA, date=NA, days=NA)
for(i in 1:dim(aggr.days)[1]){
    aggr.melt[2*i-1, ] <- aggr.days[i, ]
    aggr.melt[2*i, ] <- aggr.ends[i, ]
}

aggr.melt$days <- as.factor(aggr.melt$days)

# ploting 

library(ggplot2)

fig2 <- ggplot(data=aggr.melt, aes(x=interval, y=steps)) + geom_line()
fig2 + facet_grid(days ~ .)

```

We can see that at the weekend there are more activity in the middle of the day. Meanwhile on the workdays main activity is in the morning and in the evening. 


**In my opinion** this figure (optional) better shows workday/weekend trends

``` {r weekends2, fig.width=8, fig.height=6}
plot(
    x=aggr.days$interval, 
    y=aggr.days$steps, 
    type="l", 
    axes=F,
    xlab="time interval",
    ylab="number of steps"
)
axis(1, at=c(0,300,600,900,1200,1500,1800,2100,2400), lab=time.int)
axis(2, at=c(0,50,100,150,200))
lines(x=aggr.ends$interval, y=aggr.ends$steps, col="red")
box()
legend("topright", 
       legend=c("Weekday", "Weekend"), 
       col=1:2, 
       lty=1)
```